---
title: "Inspiration Corner"
author: "Francisco Chavez"
date: "2025-10-17"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Writer's block?  Here's the solution:

- This is an app that will help you on making progress when stuck trying to bring your next novel, all you have to do:
- The app will keep bringing up new words as you continue to write
- This will jog your memory and stimuate the thinking process!

## Instructions

- Click in the link to get to the app -> https://paco23ch.shinyapps.io/textPrediction/
- Once the page loads, follow the instructions to load the pre-trained model
- Once the file has been loaded, all you have to do is to select the settings for the model:
  - How many words to use to predict the next word
  - How many additional suggestions you want generated

## Methodology

- We used a set of data from twitter, blogs and news provided by SwiftKey
- Did a preliminary cleanup of the data, removing all offensive words, punctuation and others, to get a stream of continuous words
- We analyzed a small sample (1% of the data) of the combinations of 2, 3 and 4 words
- The idea is to identify patterns and be able to predict the next word based on the previous 1 to 3 words
- We used different sample sizes to train multiple models
- Deployed a small model as a PoC, this is only 2% of the data with a front-end to experiment.

## Algorithm

- The pre-trained model contains 4 main sets to be used (1-4 grams):
- Based on the settings, the model will use the highest preferred value to predict the next word
  - for example n-gram=4, means the model will first try to predict with the last 3 words typed in to find a suggestion
  - if no suggestions can be found, the max n-gram value will be decreased by 1 to find more options
  - if nothing can be found, then the most popular words will be used as suggestions
- The model will return the most likely word based on the probability calculated previously and the k number of values requested by the user, along with their probability

## n-gram Test results

- Sampled data to see the effect on accuracy

```{r test_results, warning=FALSE}
library(ggplot2)

test_results <- readRDS(file="../final_first/en_US/test_scores_all.rds")
dodge <- position_dodge(width = 0.9)

ggplot(test_results, aes(x = interaction(n_grams,percent), y = accuracy, fill = percent)) +
  geom_bar(stat = "identity", position = position_dodge()) + # Use geom_line for line plots, geom_point for scatter plots
  labs(title = "Accuracy for n-grams and percent sampled",
       x = "n-grams and percent sampled",
       y = "Accuracy") +
  theme_minimal()
```

